---
title: "727 HW 5"
author: "Makenna Harrison"
date: "2023-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

github: https://github.com/makharrison/727_hw5

```{r warning=FALSE,message=FALSE}
library(censusapi)
library(tidyverse)
library(magrittr)
library(factoextra)
library(ggmap)
library(ggalt)
```

```{r echo=FALSE}
cs_key = "xxxx"
```

Using censusapi, we can pull the desired data as follows:

```{r}
acs_il_c <- getCensus(name = "acs/acs5",
 vintage = 2016,
 vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"),
 region = "county:*",
 regionin = "state:17",
 key = cs_key) %>%
 rename(pop = B01003_001E,
 hh_income = B19013_001E,
 income = B19301_001E)
 head(acs_il_c)
```

Additionally, we can pull map data for counties in IL.

```{r}
 il_map <- map_data("county", region = "illinois")
 head(il_map)
```

Next, we can join the two frames together.

```{r}
acs_il_c <- acs_il_c %>%
  mutate(county_clean = tolower(NAME),
         county_clean = gsub(" county, illinois","",county_clean))
```


```{r}
acs_map <- merge(acs_il_c,il_map,by.x="county_clean",by.y="subregion")
```


```{r}
 ggplot(acs_map) +
 geom_encircle(aes(x = long, y = lat, group = group, fill = income), s_shape = 1, expand = 0,
                alpha = 0.2, color = "black", show.legend = TRUE)
```

Next, we can perform hierarchical clustering.

```{r}
acs_map_clean <- acs_map %>%
  select(pop,hh_income,income) %>%
  distinct()
```

First, we create the distance matrix

```{r}
hclust_d <- dist(acs_map_clean)
```

Next, we cluster using the ward method.

```{r}
hc_ward <- hclust(hclust_d, method = "ward.D2")
```

```{r}
plot(hc_ward, main = "Ward", xlab = "", sub = "")
rect.hclust(hc_ward, 
            k = 6, 
            border = "red")
```
There looks to be 6 clusters

```{r}
cutree(hc_ward, 6)
```

Next, we want to visualize these clusters. First, we need to add the clusters to a data frame.

```{r}
acs_map_clean2 <- acs_map %>%
  select(county_clean, pop,hh_income,income) %>%
  distinct() %>%
  mutate(cluster = cutree(hc_ward, 6)) %>%
  select(county_clean,cluster)
```

```{r}
acs_map_new <- merge(acs_map, acs_map_clean2,by="county_clean")
```


```{r}
 ggplot(acs_map_new) +
 #geom_polygon(aes(x = long, y = lat, group = group, fill = cluster)) +
 geom_encircle(aes(x = long, y = lat, group = group, fill = cluster), s_shape = 1, expand = 0,
                alpha = 0.2, color = "black", show.legend = TRUE)
```

Now, we look at census tracts instead of counties.

```{r}
acs_il_t <- getCensus(name = "acs/acs5",
vintage = 2016,
vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"),
region = "tract:*",
regionin = "state:17",
key = cs_key) %>%
mutate_all(list(~ifelse(.==-666666666, NA, .))) %>%
rename(pop = B01003_001E,
hh_income = B19013_001E,
income = B19301_001E)
head(acs_il_t)
```

Let's clean the data for kmeans clustering.

```{r}
acs_il_t_clean <- acs_il_t%>%
  select(NAME,pop,hh_income,income,county) %>%
  distinct() %>%
  na.omit() %>%
  select(pop,hh_income,income)
```

Since we want to use K Means in this section, we start by determining the optimal number
of K that results in Clusters with low within but high between variation. Plot within cluster
sums of squares for a range of K (e.g. up to 20).

The total within sum of squares is lower for larger values of k. Based on the plot, the elbow of the curve appears at k = 3.

```{r warning=FALSE}
fviz_nbclust(acs_il_t_clean, #data set we want to use
             kmeans, #cluster method
             method = "wss", #method used for estimating the optimal number of clusters
             k.max = 20)
```

```{r}
km_1 <- kmeans(acs_il_t_clean, 3, nstart = 20)
```

```{r}
acs_il_t_clean <- acs_il_t_clean %>%
  mutate(cluster = km_1$cluster)
```


The means for each cluster:

```{r}
km_1$centers
```

The most frequent counties for each cluster:

```{r}
acs_il_t_clean2 <- acs_il_t%>%
  select(NAME,pop,hh_income,income,county) %>%
  distinct() %>%
  na.omit() %>%
  mutate(cluster = acs_il_t_clean$cluster)
```

county 031 (Cook County) appears the most times for each of the three clusters.

```{r}
acs_il_t_clean2 %>% 
  group_by(cluster) %>%
  count(county) %>%
  arrange(desc(n)) %>%
  head(3) %>%
  arrange(cluster)
```

Alternatively, we could write a look for calculate kmeans for each value of k:

```{r}
clusters <- lapply(1:10, function(i) kmeans(acs_il_t_clean, i, nstart = 20))
```

Additionally, we can see which cluster each tract would fall into for differing values of k.

```{r}
acs_il_t_clean %>%
  mutate(
    cluster1 = clusters[[1]]$cluster,
    cluster2 = clusters[[2]]$cluster,
    cluster3 = clusters[[3]]$cluster,
    cluster4 = clusters[[4]]$cluster,
    cluster5 = clusters[[5]]$cluster,
    cluster6 = clusters[[6]]$cluster,
    cluster7 = clusters[[7]]$cluster,
    cluster8 = clusters[[8]]$cluster,
    cluster9 = clusters[[9]]$cluster,
    cluster10 = clusters[[10]]$cluster
  ) %>%
  head()
```

